### 2016-12-12 paper/3_gilbert_features.tex
Incorporate feedback of first reviewer
Correct latex warnings

\begin{listing}[!h]
  \begin{CenteredBox}
    \begin{lstlisting}[language=Matlab,
        commentstyle=\color{black},
        stringstyle=\color{black},
    ]
c = {true, 2*2, 'cell', 'array'};
b = c{1} & false; % = false
d = c{2} ^ 2; % = 16
s = {c{4}, c{3}}; % = {'array', 'cell'} 
    \end{lstlisting}
  \end{CenteredBox}
  \caption{Cell array usage in Gilbert. Definition of a 4 element cell array, which is accessed subsequently.}
  \label{lst:cellArray}
\end{listing}

@@

In order to terminate the fixpoint operation, the operator provides two mechanisms.
The user has to specify a maximum number \code{m} of iterations.
Additionally, the user can provide a convergence function \code{c} to the fixpoint operator.
The convergence function is called with the previous and current fixpoint value and returns a boolean value.
Thus, the fixpoint operator terminates either if convergence was detected or if the maximum number of iterations is exceeded.
\begin{equation}
fixpoint: \underbrace{T}_{\text{\code{x}}} \times \left( \underbrace{T \rightarrow T}_{\text{\code{f}}} \right) \times \underbrace{\mathbb{N}}_{\text{\code{m}}} \times \left(\underbrace{T\times T \rightarrow \mathbb{B}}_{\text{\code{c}}} \right) \rightarrow T
\label{eqn:fixpoint}
\end{equation}
with $T$ being a generic type variable.

@@

\textbf{Function Overloading}. MATLAB's basic operators, such as \code{+}, \code{-}, \code{/} and \code{*}, for example, are overloaded.
They can be applied to matrices, scalars as well as mixture of both data types.
That makes it very convenient to express mathematical problems, but from a programmer's point of view it causes some hardships.
Originally, HM cannot deal with overloaded functions properly, because it assumes that each function has an unique type.
In order to extend HM's capabilities, we allowed each function symbol to have a list of signatures.
In the case of \code{+}, the list of signatures would consist of 
\begin{eqnarray*}
matrix[double] \times matrix[double] &\rightarrow& matrix[double]\\
matrix[double] \times double &\rightarrow& matrix[double]\\
double \times matrix[double] &\rightarrow& matrix[double]\\
double \times double &\rightarrow& double
\end{eqnarray*}

@@

\textbf{Transformation Operators}. The transformation operators constitute the main abstraction of the linear algebra operations.
They group operations with similar properties and thus allow an easier reasoning and optimization of the underlying program. The \code{UnaryScalar\-Transformation} takes a single scalar value and applies an unary operation on it. The \code{ScalarScalarTransformation} constitutes a binary operation on scalar values whereas \code{ScalarMatrixTransformation} and \code{MatrixScalarTransformation} represent a binary operation between a scalar and a matrix value. The \code{VectorwiseMatrixTrans\-formation} applies an operation on each row vector of the given matrix.
A vectorwise operation produces a scalar value for each row vector. Similar to the vectorwise operations, the \code{AggregateMatrixTransformation} applies an operation to all matrix entries producing a single scalar result value. The iteration mechanism is represented by the \code{Fixpoint\-IterationMatrix} and \code{FixpointIterationCellArray} operators. The former operator is used for a fixpoint operation on matrices and the latter on cell arrays. Both operators follow the semantics and have the same parameters as the fixpoint abstraction presented in \cref{eqn:fixpoint}.

@@

\begin{listing}[!h]
  \begin{CenteredBox}
    \begin{lstlisting}[language=Matlab]
C = A'*B;
E = (C * D')';
    \end{lstlisting}
  \end{CenteredBox}
  \caption{Transpose pushdown can eliminate unnecessary transpose operations occurring in linear algebra programs.}
  \label{lst:pushdown}
\end{listing}

>>>>

\begin{listing}[!h]
    \begin{lstlisting}[language=Matlab,
        commentstyle=\color{black},
        stringstyle=\color{black},
        xleftmargin=.2\textwidth
    ]
c = {true, 2*2, 'cell', 'array'};
b = c{1} & false; % = false
d = c{2} ^ 2; % = 16
s = {c{4}, c{3}}; % = {'array', 'cell'} 
    \end{lstlisting}
  \caption{Cell array usage in Gilbert. Definition of a 4 element cell array, which is accessed subsequently.}
  \label{lst:cellArray}
\end{listing}

@@

In order to terminate the fixpoint operation, the operator provides two mechanisms.
The user has to specify a maximum number \code{m} of iterations.
Additionally, the user can provide a convergence function \code{c} to the fixpoint operator.
The convergence function is called with the previous and current fixpoint value and returns a boolean value.
Thus, the fixpoint operator terminates either if convergence was detected or if the maximum number of iterations is exceeded.
\begin{equation}
\label{eq:fixpoint}
fixpoint: \underbrace{T}_{\text{\code{x}}} \times \left( \underbrace{T \rightarrow T}_{\text{\code{f}}} \right) \times \underbrace{\mathbb{N}}_{\text{\code{m}}} \times \left(\underbrace{T\times T \rightarrow \mathbb{B}}_{\text{\code{c}}} \right) \rightarrow T
\end{equation}
with $T$ being a generic type variable.

@@

\textbf{Function Overloading}. MATLAB's basic operators, such as \code{+}, \code{-}, \code{/} and \code{*}, for example, are overloaded.
They can be applied to matrices, scalars as well as mixture of both data types.
That makes it very convenient to express mathematical problems, but from a programmer's point of view it causes some hardships.
Originally, HM cannot deal with overloaded functions properly, because it assumes that each function has a unique type.
In order to extend HM's capabilities, we allowed each function symbol to have a list of signatures.
In the case of \code{+}, the list of signatures would consist of 
\begin{alignat*}{2}
matrix[double] \times matrix[double] &\rightarrow matrix[double]\\
matrix[double] \times double &\rightarrow matrix[double]\\
double \times matrix[double] &\rightarrow matrix[double]\\
double \times double &\rightarrow double
\end{alignat*}

@@

\textbf{Transformation Operators}. The transformation operators constitute the main abstraction of the linear algebra operations.
They group operations with similar properties and thus allow an easier reasoning and optimization of the underlying program. The \code{UnaryScalar\-Transformation} takes a single scalar value and applies an unary operation on it. The \code{ScalarScalarTransformation} constitutes a binary operation on scalar values whereas \code{ScalarMatrixTransformation} and \code{MatrixScalarTransformation} represent a binary operation between a scalar and a matrix value. The \code{VectorwiseMatrixTrans\-formation} applies an operation on each row vector of the given matrix.
A vectorwise operation produces a scalar value for each row vector. Similar to the vectorwise operations, the \code{AggregateMatrixTransformation} applies an operation to all matrix entries producing a single scalar result value. The iteration mechanism is represented by the \code{Fixpoint\-IterationMatrix} and \code{FixpointIterationCellArray} operators. The former operator is used for a fixpoint operation on matrices and the latter on cell arrays. Both operators follow the semantics and have the same parameters as the fixpoint abstraction presented in \cref{eq:fixpoint}.

@@

\begin{listing}[!h]
    \begin{lstlisting}[language=Matlab, xleftmargin=.4\textwidth]
C = A'*B;
E = (C * D')';
    \end{lstlisting}
  \caption{Transpose pushdown can eliminate unnecessary transpose operations occurring in linear algebra programs.}
  \label{lst:pushdown}
\end{listing}

### 2016-12-12 paper/5_evaluation.tex
Correct latex warnings

\begin{figure}[!h]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Dimensionality $n$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				legend entries={Local, Spark, Flink},
				legend pos=north west,
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadReference};
			\addplot[
				color=red,
				mark=o,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadSpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadFlink};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				legend pos=north west,
				legend entries={Spark, Flink},
				ymin=0.0,
				ymax=2000,
				width=\dualpgfwidth,
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMult/matrixMultScalabilitySpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMult/matrixMultScalabilityFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmNodesRuntime}
	\end{subfigure}
	\caption{Scalability of matrix multiplication. \subref{fig:mmLoadRuntime} Execution time of matrix multiplication depending on the data size. \subref{fig:mmNodesRuntime} Execution time of matrix multiplication depending on the cluster size with constant work load per core.}
	\label{fig:mmScalability}
\end{figure}

@@

\begin{figure}[!h]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Number of rows of matrix $V$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				legend pos=north west,
				legend columns=2,
				legend entries={Local, Spark, Flink, HT Spark, HT Flink},
				ymax=500,
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=Rows,
				y=Time,
			]
			{data/nnmfStep/nnmfStepLoadReference};

@@

\begin{listing}[!h]
	\begin{CenteredBox}
		\begin{lstlisting}[language=Matlab,
		commentstyle=\color{black},
		  stringstyle=\color{black},
		  keywordstyle=\color{black}\bfseries,
		  morekeywords={ones, fixpoint},]
% load adjacency matrix
A = load();
maxIterations = 10;
d = sum(A, 2); % outdegree per vertex
% create the column-stochastic transition matrix
T = (diag(1 ./ d) * A)'; 
r_0 = ones(numVertices, 1) / numVertices; % initialize the ranks
e = ones(numVertices, 1) / numVertices;
% PageRank calculation
fixpoint(r_0, @(r) .85 * T * r + .15 * e, maxIterations)
		\end{lstlisting}
	\end{CenteredBox}
	\caption{Gilbert PageRank implementation.}
	\label{lst:gilbertPageRank}
\end{listing}

@@

\begin{figure}[!h]
	\centering
	\includegraphics[width=.3\linewidth]{images/pageRankStep.png}
	\caption{Data flow of one iteration of the PageRank algorithm for Spark and Flink.}
	\label{fig:pageRankDataFlow}
\end{figure}

@@

\begin{figure}[t!]
	\centering
	\begin{subfigure}[h]{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Number of vertices $n$},
				ylabel={Execution time $t$ in s},
				xmax=250000,
				ymax=50000,
				legend pos=north west,
				legend entries={Spark, Flink, SP Flink, SP Spark},
				width=\dualpgfwidth,
			]

>>>>

\begin{figure}[htbp]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Dimensionality $n$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				legend entries={Local, Spark, Flink},
				legend pos=north west,
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadReference};
			\addplot[
				color=red,
				mark=o,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadSpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMult/matrixMultLoadFlink};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				legend pos=north west,
				legend entries={Spark, Flink},
				ymin=0.0,
				ymax=2000,
				width=\dualpgfwidth,
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMult/matrixMultScalabilitySpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMult/matrixMultScalabilityFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmNodesRuntime}
	\end{subfigure}
	\caption{Scalability of matrix multiplication. \subref{fig:mmLoadRuntime} Execution time of matrix multiplication depending on the data size. \subref{fig:mmNodesRuntime} Execution time of matrix multiplication depending on the cluster size with constant work load per core.}
	\label{fig:mmScalability}
\end{figure}

@@

\begin{figure}[htbp]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Number of rows of matrix $V$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				legend pos=north west,
				legend columns=2,
				legend entries={Local, Spark, Flink, HT Spark, HT Flink},
				ymax=500,
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=Rows,
				y=Time,
			]
			{data/nnmfStep/nnmfStepLoadReference};

@@

\begin{listing}[htbp]
	\begin{lstlisting}[language=Matlab,
		commentstyle=\color{black},
		  stringstyle=\color{black},
		  keywordstyle=\color{black}\bfseries,
		  morekeywords={ones, fixpoint},
		  xleftmargin=.0\textwidth]
% load adjacency matrix
A = load();
maxIterations = 10;
d = sum(A, 2); % outdegree per vertex
% create the column-stochastic transition matrix
T = (diag(1 ./ d) * A)'; 
r_0 = ones(numVertices, 1) / numVertices; % initialize the ranks
e = ones(numVertices, 1) / numVertices;
% PageRank calculation
fixpoint(r_0, @(r) .85 * T * r + .15 * e, maxIterations)
	\end{lstlisting}
	\caption{Gilbert PageRank implementation.}
	\label{lst:gilbertPageRank}
\end{listing}

@@

\begin{figure}[htbp]
	\centering
	\includegraphics[width=.3\linewidth]{images/pageRankStep.png}
	\caption{Data flow of one iteration of the PageRank algorithm for Spark and Flink.}
	\label{fig:pageRankDataFlow}
\end{figure}

@@

\begin{figure}[htbp]
	\centering
	\begin{subfigure}[h]{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Number of vertices $n$},
				ylabel={Execution time $t$ in s},
				xmax=250000,
				ymax=50000,
				legend pos=north west,
				legend entries={Spark, Flink, SP Flink, SP Spark},
				width=\dualpgfwidth,
			]
